---
title: "Reproducible research.proj 1"
output: html_document
date: "2026-05-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the dataset 
data <- read.csv("activity.csv")

#Exploratory analysis
dim(data)
str(data)
```
```{r}
#Transform date that is as a character to date 
data$date <- as.Date(data$date)

head(data)
```

```{r}
summary(data)
```


```{r}
#Calculate total number of steps taken each day
library(dplyr)

daily_totalsteps <- data %>%
    filter(!is.na(steps)) %>% 
    group_by(date) %>%
    summarise(steps = sum(steps,na.rm=TRUE), .groups = "drop")

class(daily_totalsteps) 

# Histogram total steps/day

library(ggplot2)

ggplot(daily_totalsteps, aes(x = steps)) +
    geom_histogram(bins = 20, 
                   fill = "skyblue", 
                   color = "white", 
                   alpha = 0.85) +
    labs(
        title = "Total Number of Steps Taken Each Day",
        x = "Total number of steps",
        y = "Frequency"
    ) +
    theme_bw() +  
    theme(panel.grid.major.x = element_blank()) +
    theme(
        plot.title = element_text(          
            hjust = 0.5,                   
            size = 16,                    
            face = "bold",                 
            margin = margin(b = 15)       
        ),
        plot.subtitle = element_text(       
            hjust = 0.5,
            size = 12
        ),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.minor = element_blank()
    )
#Save graph
ggsave("plot1.png",width = 8, height = 6, dpi = 300,bg = "white")  
```
```{r}
# Calculate mean and median of the total number of steps taken per day

mean_daily_totalsteps <- mean(daily_totalsteps$steps, na.rm = TRUE)
median_daily_totalsteps <- median(daily_totalsteps$steps, na.rm = TRUE)

cat("Mean total steps per day:", round(mean_daily_totalsteps, 2), "\n") # "\n"salto de linea
cat("Median total steps per day:", round(median_daily_totalsteps, 2), "\n")
```
```{r}
# Promedio de steps por intervalo de 5 minutos
interval_avg <- aggregate(steps ~ interval, data = data, FUN = mean, na.rm = TRUE)

# Gráfico con ggplot2
ggplot(interval_avg, aes(x = interval, y = steps)) +
  geom_line(color = "darkblue", linewidth = 1) +
  labs(title = "Average Number of Steps per 5-minute Interval",
       x = "5-minute Interval",
       y = "Average Number of Steps") +
  theme_bw() +  
    theme(panel.grid.major.x = element_blank()) +
    theme(
        plot.title = element_text(          
            hjust = 0.5,                   
            size = 16,                    
            face = "bold",                 
            margin = margin(b = 15)       
        ),
        plot.subtitle = element_text(       
            hjust = 0.5,
            size = 12
        ),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.minor = element_blank()
    )
#Save graph
ggsave("plot2.png",width = 8, height = 6, dpi = 300,bg = "white")
```
```{r}
# which 5-minute interval contains the maximum number of steps?.
max_interval <- interval_avg[which.max(interval_avg$steps), ]

cat("Interval with maximum average steps:", max_interval$interval, 
    "con", round(max_interval$steps, 2), "steps\n")
```

```{r}
## Imputing missing values

### 1. Calculate and report the total number of missing values
# Total NA en el dataset
total_na <- sum(is.na(data$steps))
cat("Total number of missing values (NA):", total_na, "\n")

cat("Percentage of missing data:", round(total_na/nrow(data)*100, 2), "%\n")
```
```{r}
# Create dataset with imputed values
data_complete <- data
for (i in 1:nrow(data_complete)) {
  if (is.na(data_complete$steps[i])) {
    interval_mean <- interval_avg$steps[interval_avg$interval == data_complete$interval[i]]
   data_complete$steps[i] <- interval_mean
  }
}

# Total steps per day with imputed data
steps_per_day_complete <- aggregate(steps ~ date, data = data_complete, FUN = sum)

# Histogram of total steps per day (imputed data)
ggplot(steps_per_day_complete, aes(x = steps)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white", alpha = 0.7) +
  labs(title = "Histogram of Total Steps Taken Each Day (Imputed Data)", 
       x = "Total Steps", y = "Frequency") +
 theme_bw() +  
    theme(panel.grid.major.x = element_blank()) +
    theme(
        plot.title = element_text(          
            hjust = 0.5,                   
            size = 16,                    
            face = "bold",                 
            margin = margin(b = 15)       
        ),
        plot.subtitle = element_text(       
            hjust = 0.5,
            size = 12
        ),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.minor = element_blank()
    )
#Save graph
ggsave("plot3.png", width = 8, height = 6, dpi = 300,bg = "white") 
```


```{r}
# Calculate mean and median of the total number of steps taken per day

mean_steps_per_day_complete <- mean(steps_per_day_complete$steps, na.rm = TRUE)
median_steps_per_day_complete <- median(steps_per_day_complete$steps, na.rm = TRUE)

cat("Mean total steps per day:", round(mean_steps_per_day_complete, 2), "\n") # "\n"salto de linea
cat("Median total steps per day:", round(median_steps_per_day_complete, 2), "\n")
```

Impact of imputation: The mean remains the same (as expected since we used interval means), but the median increased slightly from r round(median_steps, 2) to r round(median_steps_complete, 2). The histogram shows a slight shift toward higher step counts.


```{r}
# Create factor variable for weekday vs weekend

data_complete$day_type <- ifelse(weekdays(data_complete$date) %in% c("Saturday", "Sunday"),"weekend", "weekday")

data_complete$day_type <- as.factor(data_complete$day_type)

# Average steps per interval by day type
avg_steps_daytype <- aggregate(steps ~ interval + day_type, data = data_complete, FUN = mean)

# Panel plot
ggplot(avg_steps_daytype, aes(x = interval, y = steps, color = day_type)) +
  geom_line(linewidth = 0.6, alpha = 0.9) +
  facet_wrap(~ day_type, ncol = 1, scales = "free_y") +
  scale_color_manual(
    values = c("weekday" = "steelblue",    
               "weekend" = "magenta")) +
  scale_y_continuous(breaks = seq(0, 250, 50), limits = c(0, 250)) + 
  labs(title = "Average Steps: Weekdays vs Weekends",
    subtitle = "5-minute intervals",
    x = "5-minute Interval",
    y = "Average Steps") +
  theme_bw() +  
    theme(panel.grid.major.x = element_blank()) +
    theme(
        plot.title = element_text(          
            hjust = 0.5,                   
            size = 16,                    
            face = "bold",                 
            margin = margin(b = 15)),
        plot.subtitle = element_text(       
            hjust = 0.5,
            size = 12),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.minor = element_blank()
    )

#Save graph
ggsave("plot4.png", width = 8, height = 6, dpi = 300,bg = "white") 
```
WEEKDAYS vs WEEKENDS

WEEKDAYS (61 observations):
High peaks: intervals 500-1000 and 1500-2000 
Maximum: 206 steps (interval 835)
More intense but concentrated activity

WEEKENDS (42 observations): 
More uniform activity throughout the day
Maximum: ~180 steps
No pronounced peaks
Greater sustained morning activity
KEY INSIGHTS FROM FIXED Y-SCALE COMPARISON:








